A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-09-08 bioc-workflow.Rmd
Make pretty for Laurent
added filterBinMSnSet

```{r runPD, eval = FALSE}
## As per Christoforou et al (2016), 
hl <- phenoDisco(hl, fcol = "phenoDisco.Input", 
                 times = 200, 
                 GS = 20)
```

@@

```{r makego, eval=FALSE}
gocc <- makeGoSet(hl, params = par, 
                  namespace = "cellular_component", 
                  evidence = NULL)
gocc
```

@@

The first step, as with any machine learning algorithm, is to optimise
any free paramaters of the classifier. For the *k*-NN TL classifier
there are two sets of parameters that need optimising: the first set
are the $k$'s for the primary and auxiliary data sources required for
the nearest neighbour calculations for each data source.  The second
set of parameters (which is noted by a vector of $\theta$ weights)
that require optimising are the class weights, one per subcellular
niche, that control the proportion of primary and auxiliary data to
use for learning. A weight can take any real value number between 0
and 1. A weight of $\theta = 1$ indicates that all weight is given to
the primary data (and this implicitly implies that a weight of $1 -
\theta$ is given to the auxiliary data, similarly a weight of $\theta
= 0$ implies that all weight is given to the auxiliary data (so 0 is
given to the primary source). If we conduct a parameter seach and test
weights $\theta = {0, 1/3, 2/3, 1}$ for each class, and if we have,
for example 10 subcellular niches, this will result in `4^10`
different combinations of parameters to test. The parameter
optimisation is therefore time consuming and as such we recommend you
make use of a computing cluster (code and submissing scripts are also
available in the supporting information). The markers in the `hl`
dataset contain 14 subcellular classes. If we examine these markers
and classes on the PCA plot above we can see that in particular the
two ribosomes and two nuclear compartments are highly separated along
the first two components, this is also evident from the profiles plot
which gives us a good indication that these subcellular niches are
well-resolved in the hyperLOPIT dataset. Transfer learning is
particularly useful for classes that are not as well separated, we
find that subcellular niches that are well-separated under hyperLOPIT
and LOPIT obtain a class score of 1 (i.e. use only primary data from
transfer learning [@Breckels:2016]). Therefore, for the optimisation
stage of the analyses we can already infer a subcellular class weight
of 1 for these niches and only optimise over the remaining organelles.
This can signifciantly cut down optimisation time as by removing these
4 classes from the optimisation (and not the classification) we only
have `4^10` class weight combinations to consider instead of `4^14`
combinations. In the example below we remove these 4 classes from the
marker set, re-run the `knnOptimisation` for each data source and then
run the `knntlOptimisation` with the 10 remaining classes. (Note: this
is not run live as this the `hl` dataset with 10 classes, 707 markers
and `4^10` combinations of parameters takes ~76 hours to run on the
University of Cambridge HPC using 256 workers).

@@

```{r loadtlres, include = FALSE}
load("Data/tl_runs/tlopt.rda")
data("hyperLOPIT2015goCC")
gocc <-hyperLOPIT2015goCC
ka <- kp <- 3
```
```{r setuptlopt, eval=TRUE}
## create new markers column for tl markers
fData(hl)$tlmarkers <- fData(hl)$markers
fData(gocc)$tlmarkers <- fData(gocc)$markers

>>>>

```{r runPD, eval = FALSE}
## As per Christoforou et al (2016), 
hl <- phenoDisco(hl, fcol = "phenoDisco.Input", 
                 times = 200, 
                 GS = 60)
```

@@

```{r makego, eval=FALSE}
gocc <- makeGoSet(hl, params = par, 
                  namespace = "cellular_component", 
                  evidence = NULL)
```

The function `makeGoSet` uses the `r Biocpkg("biomaRt")` package to query the relevent database (e.g. Ensembl, Uniprot) for GO terms. All GO terms that have been observed for the 5032 proteins in the hyperLOPIT dataset are retieved. Users should note that the number of GO terms retreived is also dependent on the database version queried and thus is always subject to change. We find it is common to see many GO terms with only one protein assigned to that term, these such terms generally do not bring and information generally for building the classifier and thus we can remove such GO terms by using the function `filterBinMSnSet`.

```{r filtergo, eval=FALSE}
gocc <- filterBinMSnSet(hl)
```

@@

The first step, as with any machine learning algorithm, is to optimise
any free paramaters of the classifier. For the *k*-NN TL classifier
there are two sets of parameters that need optimising: the first set
are the $k$'s for the primary and auxiliary data sources required for
the nearest neighbour calculations for each data source.

```{r knnoptfortl, eval=FALSE}
## get best k's
kpopt <- knnOptimisation(hl, fcol = "tlmarkers")
kaopt <- knnOptimisation(gocc, fcol = "tlmarkers")
kp <- getParams(kpopt)
ka <- getParams(kaopt)
```

The second
set of parameters (which is noted by a vector of $\theta$ weights)
that require optimising are the class weights, one per subcellular
niche, that control the proportion of primary and auxiliary data to
use for learning. A weight can take any real value number between 0
and 1. A weight of $\theta = 1$ indicates that all weight is given to
the primary data (and this implicitly implies that a weight of $1 -
\theta$ is given to the auxiliary data, similarly a weight of $\theta
= 0$ implies that all weight is given to the auxiliary data (so 0 is
given to the primary source). If we conduct a parameter seach and test
weights $\theta = {0, 1/3, 2/3, 1}$ for each class, and if we have,
for example 10 subcellular niches, this will result in `4^10`
different combinations of parameters to test. The parameter
optimisation is therefore time consuming and as such we recommend you
make use of a computing cluster (code and submissing scripts are also
available in the supporting information). The markers in the `hl`
dataset contain 14 subcellular classes. If we examine these markers
and classes on the PCA plot above we can see that in particular the
two ribosomes and two nuclear compartments are highly separated along
the first two components, this is also evident from the profiles plot
which gives us a good indication that these subcellular niches are
well-resolved in the hyperLOPIT dataset. Transfer learning is
particularly useful for classes that are not as well separated, we
find that subcellular niches that are well-separated under hyperLOPIT
and LOPIT obtain a class score of 1 (i.e. use only primary data from
transfer learning [@Breckels:2016]). Therefore, for the optimisation
stage of the analyses we can already infer a subcellular class weight
of 1 for these niches and only optimise over the remaining organelles.
This can signifciantly cut down optimisation time as by removing these
4 classes from the optimisation (and not the classification) we only
have `4^10` class weight combinations to consider instead of `4^14`
combinations. In the example below we remove these 4 classes from the
marker set, re-run the `knnOptimisation` for each data source and then
run the `knntlOptimisation` with the 10 remaining classes. (Note: this
is not run live as this the `hl` dataset with 10 classes, 707 markers
and `4^10` combinations of parameters takes ~76 hours to run on the
University of Cambridge HPC using 256 workers).

@@

```{r loadtlres, include = FALSE}
load("Data/tl_runs/tlopt.rda")
data("hyperLOPIT2015goCC")
gocc <-hyperLOPIT2015goCC
ka <- kp <- 3
```

```{r setuptlopt, eval=TRUE}
## create new markers column for tl markers
fData(hl)$tlmarkers <- fData(hl)$markers
fData(gocc)$tlmarkers <- fData(gocc)$markers
